---
title: "Warm Up Assignment"
author: "Diana Lin"
date: "09/01/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## README

The README for this _repository_ can be found [here](../README.md).

## Exploring the `Titanic` Dataset
Before we start, let's load some packages that might be helpful:

```{r load packages, message = FALSE, warning = FALSE}
library(tidyverse) # for dplyr and ggplot
```

First, let's convert the `Titanic` dataset into a dataframe, and then take a brief glimpse into it.
```{r load data}
titanic_df <- data.frame(Titanic) # convert the dataset into a dataframe
summary(titanic_df) # glimpse at summary of the dataset
```

### Passenger Breakdown

#### Children vs. Adults
To find out how many children and adults were on the Titanic, we will use the `dplyr` function `filter()` and `select()`, and the `baseR` function `sum()`:

```{r children vs adults}
# Count number of adults
num_adults <- titanic_df %>% 
  # Input is the titanic dataframe
  filter(Age == "Adult") %>% # filter for rows where the passenger is an Adult
  select(Freq) %>% # Select the frequency column
  sum()

# Count number of children
num_children <- titanic_df %>%
  filter(Age == "Child") %>% # filter for rows where the passenger is a Child
  select(Freq) %>%
  sum() 
```

**Conclusion**: There were `r num_children` children and `r num_adults` adults onboard the Titanic.

#### Male vs. Female Adults
Next, we want to find out if there were more female adult or male adult passengers. To do this, let's do the same thing as above but with the `Sex`  and `Age` column.
```{r adult females vs males}
# Count number of females
num_females <- titanic_df %>%
  filter(Sex == "Female", Age == "Adult") %>%
  select(Freq) %>%
  sum()

# Count number of males
num_males <- titanic_df %>%
  filter(Sex == "Male", Age == "Adult") %>%
  select(Freq) %>%
  sum()

# case statement for the inline R code conclusion below
sex <- case_when(
  num_females > num_males ~ "more female adult passengers",
  num_females < num_males ~ "more male adult passengers",
  num_females == num_males ~ "the same number of male and female adult passengers"
)
```

There were `r num_females` female adults and `r num_males` male adults onboard the Titanic.

**Conclusion**: There were `r sex` on the Titanic.

### Survival Rates

#### Adult vs. Child

To find out whether the children had better survival rate than adults, we will need to check the columns `Age` and `Survived`, filtering and counting. Then we will take that number and divide it by the number in each age group that boarded the Titanic.

```{r age survival}
# Count the number of children survived
num_children_survived <- titanic_df %>%
  filter(Age == "Child", Survived == "Yes") %>%
  select(Freq) %>%
  sum()

# Count the numer of adults survived
num_adults_survived <- titanic_df %>%
  filter(Age == "Adult", Survived == "Yes") %>%
  select(Freq) %>%
  sum()

# Calculate the survival rate for children
children_survival <- round(num_children_survived / num_children*100, digits = 1)

# Calculate the survival rate for adults
adult_survival <- round(num_adults_survived / num_adults*100, digits = 1)

# case statement for inline R code conclusion
better_survival <- case_when(
  children_survival > adult_survival ~ "children had a better survival rate",
  adult_survival > children_survival ~ "adults had a better survival rate",
  adult_survival == children_survival ~ "adults and children had the same survival rate"
)
```

`r num_children_survived` out of `r num_children` child passengers survived.

`r num_adults_survived` out of `r num_adults` adult passengers survived.

Age | Survival Rate
----|---------------
Child | `r children_survival`%
Adult | `r adult_survival`%

**Conclusion**: The `r better_survival` on the Titanic.

#### First vs. Second vs. Third vs. Crew Class

Next, we look at whether the class of the passengers impacted the survival rate. To do this, we repeat the above, but by looking at `Class` and `Survived`.

##### First Class
Let's look at first class:
```{r first class survival}
# Count number of first class passengers that boarded the Titanic
num_first <- titanic_df %>% 
  filter(Class == "1st") %>%
  select(Freq) %>%
  sum()

# Count number of survived first class passengers
num_first_survived <- titanic_df %>% 
  filter(Class == "1st", Survived == "Yes") %>%
  select(Freq) %>%
  sum()

# Calculate the survival rate
first_survival <- num_first_survived / num_first

percent_first <- round(first_survival*100,digits = 1)
```

`r num_first_survived` out of `r num_first` passengers in first class survived.

First class had a survival rate of `r percent_first`%.

##### Second Class

Next, let's look at second class:
```{r second class survival}
# Count number of second class passengers that boarded the Titanic
num_second <- titanic_df %>% 
  filter(Class == "2nd") %>%
  select(Freq) %>%
  sum()

# Count number of survived second class passengers
num_second_survived <- titanic_df %>% 
  filter(Class == "2nd", Survived == "Yes") %>%
  select(Freq) %>%
  sum()

# Calculate the survival rate
second_survival <- num_second_survived / num_second
percent_second <- round(second_survival*100, digits=1)
```

`r num_second_survived` out of `r num_second` passengers in second class survived.

Second class had a survival rate of `r percent_second`%.

##### Third Class
Now, let's look at third class:
```{r third class survival}
# Count number of third class passengers that boarded the Titanic
num_third <- titanic_df %>% 
  filter(Class == "3rd") %>%
  select(Freq) %>%
  sum()

# Count number of survived third class passengers
num_third_survived <- titanic_df %>% 
  filter(Class == "3rd", Survived == "Yes") %>%
  select(Freq) %>%
  sum()

# Calculate the survival rate
third_survival <- num_third_survived / num_third
percent_third <- round(third_survival*100, digits=1)
```

`r num_third_survived` out of `r num_third` passengers in third class survived.

Third class had a survival rate of `r percent_third`%.

##### Crew

Lastly, let's look at the crew:
```{r crew survival}
# Count number of crew  that boarded the Titanic
num_crew <- titanic_df %>% 
  filter(Class == "Crew") %>%
  select(Freq) %>%
  sum()

# Count number of survived crew
num_crew_survived <- titanic_df %>% 
  filter(Class == "Crew", Survived == "Yes") %>%
  select(Freq) %>%
  sum()

# Calculate the survival rate
crew_survival <- num_crew_survived / num_crew
percent_crew <- round(crew_survival*100, digits=1)
```

`r num_crew_survived` out of `r num_crew` crew members survived.

Crew members had a survival rate of `r percent_crew`%.

##### Summary
Let's summarize our survival rates so far:

Class | Survival Rate
------|---------------
First | `r percent_first`%
Second | `r percent_second`%
Third | `r percent_third`%
Crew | `r percent_crew`%

Now to computationally compare these survival rates and return the class with the best survival rate:
```{r best survival}
# put all survival rates into a vector
class_survival <- c(first_survival, second_survival, third_survival, crew_survival)

# have a corresponding name for each survival rate
class_survival_name <- c("first", "second", "third", "crew")

# find the highest value in the vector, best survival rate value
best_survival <- max(class_survival)

# find the corresponding indices of the vector that correspond to the max value and then get the name corresponding
best_survival_class <-  class_survival_name[which(class_survival %in% best_survival)]

# if all are the same survival rate, have a different statement/conclusion
best_survival_class_final <- if_else(
  var(class_survival) == 0, 
  "four classes had the same survival rate",
  paste(paste(best_survival_class, collapse = ', '),
        ifelse(
          length(best_survival_class) > 1,
          "classes had the best survival rate" ,
          "class had the best survival rate")))
```

**Conclusion**: The `r best_survival_class_final` on the Titanic.

## Visualizing `ToothGrowth`

### Data
First, let's read in the file:
```{r load toothgrowth}
toothgrowth <- read.table("https://raw.githubusercontent.com/STAT540-UBC/STAT540-UBC.github.io/master/homework/practice_assignment/guinea_pigs_tooth_growth.txt", header = TRUE)
```

### Summary
Let's take a glimpse at the summary:
```{r glimpse toothgrowth}
summary(toothgrowth)
```

### Plot
Let's plot a mix of a scatter plot and a violin plot:

```{r plot-toothgrowth}
toothgrowth %>%
  ggplot(aes(x=factor(dose), y=len, fill=supp)) +
  geom_violin(position = position_dodge(width = 0.75)) + # use width = 0.75 since the default for position_jitterdodge() below default width is 0.75, factor since dose is a 'category'
  geom_point(position = position_jitterdodge()) + # use jitter to scatter the plots so that they're not all in one line, easier to distinguish points that would normally overalp
  labs(title = "The Impact of Vitamin C Doses on Guinea Pig Tooth Growth", 
       subtitle = "Delivery Methods: Orange Juice (OJ) vs Ascorbic Acid (VC)",
       x = "Dose (mg/day)",
       y = "Tooth Length (units not given)",
       fill = "Delivery\nMethod"
  ) +
  theme_bw() +
  theme(text = element_text(size=12))
```

### Observations
What does this plot show us?

1. Tooth length increases as the dose increases.
1. For the doses of 0.5 mg/day and 1.0 mg/day, orange juice seems to be the more effective delivery method of Vitamin C, promoting more tooth growth (increase in tooth length) than ascorbic acid.
1. For the dose of 2.0 mg/day, neither orange juice nor ascorbic acid seems to be more effective than the other.

### Plot Rationale
I chose this combination of a violin plot and a scatter plot for 2 reasons:

1. **Violin plot**: to see the density/distribution of each data point.
1. **Scatter plot**: there aren't enough data points in this dataset for a violin plot alone to be very informative-- since there are only a few points, being able to see each individual point is better.