---
title: "Seminar 02b: Graphing using ggplot2"
author: "Diana Lin"
date: "14/01/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading the tidyverse package:
```{r load tidyverse, message=FALSE, warning=FALSE}
library(tidyverse)
```

# Deliverable

Here is the deliverable for seminar 02b:
```{r deliverable}
mpg %>%
  ggplot(aes(x=displ, y=hwy, size=class, colour=drv)) +
  geom_point()
```

# Seminar Material

To see what the `mpg` dataset looks like:
```{r mpg}
mpg
```

To plot `mpg`:
```{r plot}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

To add a third variable:
```{r third}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

Trying out `size` as the third variable:
```{r size}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

To change the colour of ALL points regardless of any properties:
```{r colour}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

To see all the available `geom` functions:
```{r geom}
apropos("^geom_")
```

To see all the available `stat` functions:
```{r stat}
apropos("^stat_")
```

To see all the available `scale` functions:
```{r scale}
apropos("^scale_")
```

To add a smooth line (regression with loess) for `engine size` vs `fuel efficiency`:
```{r smooth-line}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth()
```

Adding the mapping to both `geom_point` and `geom_smooth` will result in the same plot as having it in `ggplot`:

```{r smooth-line-2}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

To add a continuous variable as the third dimension:
```{r third-dimension}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = hwy, color = year)) +
  geom_point() +
  geom_smooth()
```

What happens if you add a third discrete variable?
```{r third-dimension-2}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = hwy, color = trans)) +
  geom_point() +
  geom_smooth()
```

Finding average fuel efficiency for each class:
```{r data}
(averageEfficiency <- 
  mpg %>% group_by(class) %>% summarise(fuel_efficiency = mean(hwy)))
```

To use `geom_bar` with specific heights, use `stat = identity`:
```{r bar}
ggplot(averageEfficiency) + 
  geom_bar(aes(x = class, y = fuel_efficiency),
           stat = "identity")
```

To add colour to the bar chart:
```{r bar-colour}
ggplot(averageEfficiency) + 
  geom_bar(aes(x = class, y = fuel_efficiency, fill = class),
           stat = "identity")
```

To specify the x-axis and y-axis labels:
```{r axis-label}
ggplot(averageEfficiency) + 
  geom_bar(aes(x = class, y = fuel_efficiency, fill = class),
           stat = "identity") +
  ylab("Fuel Efficiency (miles per gallon)") +
  xlab("Vehicle Type")
```

To reverse the y-scale:
```{r yscale}
ggplot(averageEfficiency) + 
  geom_bar(aes(x = class, y = fuel_efficiency, fill = class),
           stat = "identity") +
  ylab("Fuel Efficiency (miles per gallon)") +
  xlab("Vehicle Type") +
  scale_y_reverse()
```

To add to the coordinate system default `coord_cartesian` with `coord_flip`:

```{r coord}
ggplot(averageEfficiency) + 
  geom_bar(aes(x = class, y = fuel_efficiency, fill = class),
           stat = "identity") +
  coord_flip()
```

How about `coord_polar`?
```{r polar}
ggplot(averageEfficiency) + 
  geom_bar(aes(x = class, y = fuel_efficiency, fill = class),
           stat = "identity") +
  coord_polar()
```

To facet the original scatter plot by `class`:

```{r facet}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~class)
```

