---
title: "seminar06"
author: "Diana Lin"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE)
```
## Deliverables

1. [Heatmap](#heatmap)
2. [Dendrogram](#dendrogram)
3. [PCA](#pca)

## Setup

Load the libraries:
```{r load pkg}
library(RColorBrewer)
library(cluster)
library(pvclust)
library(xtable)
library(limma)
library(plyr)
library(lattice)
library(RCurl)
options(download.file.method = "curl")
library(GEOquery)
library(knitr)
library(pheatmap)
library(here)
library(ggplot2)
```

Download the data:
```{r download data}
if (file.exists(here::here("seminars", "seminar06", "GSE70213.Rdata"))) {
  # if previously downloaded
  load(here::here("seminars", "seminar06", "GSE70213.Rdata"))
} else {
  # Get geo object that contains our data and phenotype information
  geo_obj <- getGEO("GSE70213", GSEMatrix = TRUE)
  geo_obj <- geo_obj[[1]]
  save(geo_obj,
       file = here::here("seminars", "seminar06", "GSE70213.Rdata"))
}
```

Getting the expression data:
```{r load data}
data <- exprs(geo_obj)
```

Getting the covariate data:
```{r covariate}
prDes <-
  pData(geo_obj)[, c("organism_ch1", "title", colnames(pData(geo_obj))[grep("characteristics",
                                                                            colnames(pData(geo_obj)))])]
```

Cleaning the covariate data:
```{r clean}
colnames(prDes) = c("organism", "sample_name", "tissue", "genotype", "sex", "age")
prDes$tissue = as.factor(gsub("tissue: ", "", prDes$tissue))
prDes$genotype = as.factor(gsub("genotype: ", "", prDes$genotype))
prDes$sex = as.factor(gsub("Sex: ", "", prDes$sex))
prDes$age = gsub("age: ", "", prDes$age)
```

## Exploratory Analysis

```{r look data}
kable(head(data[, 1:5]))
dim(data)
kable(head(prDes))
dim(prDes)
```

Frequency histogram to see how the gene values are spread across our dataset:
```{r hist}
hist(data, col = "gray", main = "GSE70213 - Histogram")
```

Since there are a lot of values < 1000, we use a log~2~ scale to transform the data:

```{r log hist}
hist(log2(data + 1), col = "gray", main = "GSE70213 log transformed - Histogram")
```

Rescale the rows in our data object:
```{r rescale}
sprDat <- t(scale(t(data)))
str(sprDat, max.level = 0, give.attr = FALSE)
round(
  data.frame(
    avgBefore = rowMeans(head(data)),
    avgAfter = rowMeans(head(sprDat)),
    varBefore = apply(head(data), 1, var),
    varAfter = apply(head(sprDat), 1, var)
  ),
  2
)
```

Now the data for each row (one probeset) has a mean of 0 and variance of 1.

## Sample Clustering

1. Cluster the data using agglomerative hierarchical clustering.
    - partitions can be visualized using a **dendrogram** at various levels of granularity -- no need to input the number of clusters
2. Find various clustering solutions using partitional clustering methods, specifically K-means and partition around medoids (PAM).
    - partitions are independent of each other, and the number of clusters is given as an input

### Part I: Hierarchical Clustering Methods

#### Hierarchical Clustering for Mice Knockout Data

```{r dend}
data_to_plot = sprDat
# compute pairwise distances
pr.dis <- dist(t(data_to_plot), method = "euclidean")

# create a new factor representing the interaction of tissue type and genotype
prDes$grp <- with(prDes, interaction(tissue, genotype))
summary(prDes$grp)

# compute hierarchical clustering using different linkage types
pr.hc.s <- hclust(pr.dis, method = "single")
pr.hc.c <- hclust(pr.dis, method = "complete")
pr.hc.a <- hclust(pr.dis, method = "average")
pr.hc.w <- hclust(pr.dis, method = "ward.D")

# plot them
op <- par(mar = c(0, 4, 4, 2), mfrow = c(2, 2))

plot(pr.hc.s, labels = FALSE, main = "Single", xlab = "")
plot(pr.hc.c, labels = FALSE, main = "Complete", xlab = "")
plot(pr.hc.a, labels = FALSE, main = "Average", xlab = "")
plot(pr.hc.w, labels = FALSE, main = "Ward", xlab = "")

par(op)
```

<!-- > **Exercise:** Play with the options of the pheatmap function and compare the different heatmaps. Note that one can also use the original data data and set the option scale = "row". You will get the same heatmaps although the columns may be ordered differently (use cluster_cols = FALSE to suppress reordering). -->

```{r heatmap}
# set pheatmap clustering parameters
clust_dist_col = "euclidean"  #‘'correlation'’ for Pearson correlation, ‘'euclidean'’, ‘'maximum'’, ‘'manhattan'’, ‘'canberra'’, ‘'binary'’ or ‘'minkowski'’
clust_method = "ward.D2"  #‘'ward.D'’, ‘'ward.D2'’,‘'single'’, ‘'complete'’, ‘'average'’ (= UPGMA), ‘'mcquitty'’ (= WPGMA), ‘'median'’ (= WPGMC) or ‘'centroid'’ (= UPGMC)
clust_scale = "none"  #'column', 'none', 'row'

## the annotation option uses the covariate object (prDes) we defined. It should
## have the same rownames, as the colnames in our data object (data_to_plot).

pheatmap(
  data_to_plot,
  cluster_rows = FALSE,
  scale = clust_scale,
  clustering_method = clust_method,
  clustering_distance_cols = clust_dist_col,
  show_colnames = T,
  show_rownames = FALSE,
  main = "Clustering heatmap for GSE70213",
  annotation = prDes[, c("tissue", "genotype",
                         "grp")]
)
```

Change the colours of the different covariates:
```{r colour}
## We can change the colours of the covariates
var1 = c("orange1", "darkred")
names(var1) = levels(prDes$tissue)
var2 = c("grey", "black")
names(var2) = levels(prDes$genotype)
var3 = c("pink1", "pink3", "lightblue1", "blue3")
names(var3) = levels(as.factor(prDes$grp))
covar_color = list(tissue = var1,
                   genotype = var2,
                   grp = var3)

my_heatmap_obj = pheatmap(
  data_to_plot,
  cluster_rows = FALSE,
  scale = clust_scale,
  clustering_method = clust_method,
  clustering_distance_cols = clust_dist_col,
  show_rownames = FALSE,
  main = "Clustering heatmap for GSE70213",
  annotation = prDes[,
                     c("tissue", "genotype", "grp")],
  annotation_colors = covar_color
)
```

Get clusters from `pheatmap` object:
```{r cluster heatmap}
cluster_samples = cutree(my_heatmap_obj$tree_col, k = 10)
# cluster_genes = cutree(my_heatmap_obj$tree_row, k=100)
kable(cluster_samples)
```

Do the same but using `hclust`:

```{r hclust}
# identify 10 clusters
op <- par(mar = c(1, 4, 4, 1))
plot(pr.hc.w, labels = prDes$grp, cex = 0.6, main = "Ward showing 10 clusters")
rect.hclust(pr.hc.w, k = 10)
par(op)
```

Saving the heatmap:

```{r save heatmap}
# Save the heatmap to a PDF file
pdf("GSE70213_Heatmap.pdf")
pheatmap(data_to_plot, cluster_rows = F, scale = clust_scale, clustering_method = clust_method, 
    clustering_distance_cols = clust_dist_col, annotation = prDes[, c("tissue", "genotype", 
        "grp")], annotation_colors = covar_color)
dev.off()
```

### Part II: Parametric and Alternative Non-Parametric Clustering with PCA and t-SNE

#### Partitioning Methods for Mice Knockout Data

K-means clustering of samples using all genes:
```{r kmeans}
# Objects in columns
set.seed(31)
k <- 5
pr.km <- kmeans(t(data_to_plot), centers = k, nstart = 50)

# We can look at the within sum of squares of each cluster
pr.km$withinss

# We can look at the composition of each cluster
pr.kmTable <- data.frame(exptStage = prDes$grp, cluster = pr.km$cluster)
kable(pr.kmTable)
```

Using the PAM algorithmn, to identify a suitable k:
```{r pam}
pr.pam <- pam(pr.dis, k = k)
pr.pamTable <- data.frame(exptStage = prDes$grp, cluster = pr.pam$clustering)
kable(pr.pamTable)
```

Making a silhouette plot:
```{r silhouette}
op <- par(mar = c(5, 1, 4, 4))
plot(pr.pam, main = "Silhouette Plot for 5 clusters")
par(op)
```

<!--> Take-home problem (1) Draw a plot with number of clusters in the x-axis and the average silhouette widths in the y-axis. Use the information obtained to determine if 5 was the best choice for the number of clusters.

> Take-home problem (2) For a common choice of k, compare the clustering across different methods, e.g. hierarchical (pruned to specific k, obviously), k-means, PAM. You will re-discover the "label switching problem" for yourself. How does that manifest itself? How concordant are the clusterings for different methods?
-->

## Gene Clustering

### A Smaller Dataset

Using `limma`:
```{r limma}
cutoff <- 1e-05
DesMat <- model.matrix(~grp, prDes)
dsFit <- lmFit(sprDat, DesMat)
dsEbFit <- eBayes(dsFit)
dsHits <- topTable(dsEbFit, coef = grep("grp", colnames(coef(dsEbFit))), p.value = cutoff, 
    n = Inf)
numBHhits <- nrow(dsHits)

topGenes <- rownames(dsHits)

# Scaled data of topGenes
topDat <- sprDat[topGenes, ]
```

### Agglomerative Hierarchical Clustering

Plot the heatmap using the `pheatmap` function:
```{r plot heatmap}
pheatmap(topDat, cluster_rows = TRUE, scale = "none", clustering_method = "average", 
    clustering_distance_cols = "euclidean", clustering_distance_rows = "euclidean", 
    annotation = prDes[, c("tissue", "genotype", "grp")], show_rownames = FALSE, 
    annotation_colors = covar_color)
```

Plot the heatmap using the `plot` function after making the `hclust` object:
```{r heatmap hclust}
geneC.dis <- dist(topDat, method = "euclidean")

geneC.hc.a <- hclust(geneC.dis, method = "average")

plot(geneC.hc.a, labels = FALSE, main = "Hierarchical with Average Linkage", xlab = "")
```

### Partitioning Methods

Visualize a cluster using line plots:
```{r partition}
set.seed(1234)
k <- 5
kmeans.genes <- kmeans(topDat, centers = k)

# choose which cluster we want
clusterNum <- 2

# Set up the axes without plotting; ylim set based on trial run.
plot(kmeans.genes$centers[clusterNum, ], ylim = c(0, 10), type = "n", xlab = "Samples", 
    ylab = "Relative expression")

# Plot the expression of all the genes in the selected cluster in grey.
matlines(y = t(topDat[kmeans.genes$cluster == clusterNum, ]), col = "grey")

# Add the cluster center. This is last so it isn't underneath the members
points(kmeans.genes$centers[clusterNum, ], type = "l")

# Optional: colored points to show which stage the samples are from.
points(kmeans.genes$centers[clusterNum, ], col = prDes$grp, pch = 20)
```

## Evaluating Clusters

### Choosing the Right k

Calculating the AIC for each choice of k:
```{r calc k}
set.seed(31)

k_max <- 10  # the max number of clusters to explore clustering with 
km_fit <- list()  # create empty list to store the kmeans object

for (i in 1:k_max) {
    k_cluster <- kmeans(t(sprDat), centers = i, nstart = 50)
    km_fit[[i]] <- k_cluster
}


# calculate AIC
km_AIC <- function(km_cluster) {
    m <- ncol(km_cluster$centers)
    n <- length(km_cluster$cluster)
    k <- nrow(km_cluster$centers)
    D <- km_cluster$tot.withinss
    return(D + 2 * m * k)
}
```

Plot the AIC vs the number of clusters:
```{r aic vs clust}
aic <- sapply(km_fit, km_AIC)
plot(seq(1, k_max), aic, xlab = "Number of clusters", ylab = "AIC", pch = 20, cex = 2, 
    main = "Clustering Samples")
```

Do the same for BIC:
```{r bic}
# calculate BIC
km_BIC <- function(km_cluster) {
    m <- ncol(km_cluster$centers)
    n <- length(km_cluster$cluster)
    k <- nrow(km_cluster$centers)
    D <- km_cluster$tot.withinss
    return(D + log(n) * m * k)
}

bic <- sapply(km_fit, km_BIC)
plot(seq(1, k_max), bic, xlab = "Number of clusters", ylab = "BIC", pch = 20, cex = 2, 
    main = "Clustering Samples")
```

### Statistical Methods

Identify the most stable clustering arising from hierarchical clustering:
```{r bootstrap}
pvc <- pvclust(topDat, nboot = 100)
plot(pvc, labels = prDes$grp, cex = 0.6)
pvrect(pvc, alpha = 0.95)
```

## Feature Reduction

### PCA Plots

Plot a PCA plot using `prcomp()` (or `svd()`):
```{r pca}
pcs <- prcomp(sprDat, center = FALSE, scale = FALSE)

# scree plot
plot(pcs)

# append the rotations for the first 10 PCs to the phenodata
prinComp <- cbind(prDes, pcs$rotation[rownames(prDes), 1:10])

# scatter plot showing us how the first few PCs relate to covariates
plot(prinComp[, c("genotype", "tissue", "PC1", "PC2", "PC3")], pch = 19, cex = 0.8)
```

Spread of the samples, explained by their 2 principal components:
```{r spread}
plot(prinComp[, c("PC1", "PC2")], pch = 21, cex = 1.5)
```

Localization of the covariate `tissue` in the different clusters:
```{r tissue}
plot(prinComp[, c("PC1", "PC2")], bg = prDes$tissue, pch = 21, cex = 1.5)
legend(list(x = 0.2, y = 0.3), as.character(levels(prDes$tissue)), pch = 21, pt.bg = c(1, 
    2, 3, 4, 5))
```

Localization of the covariate `genotype` in the different clusters:
```{r genotype}
plot(prinComp[, c("PC1", "PC2")], bg = prDes$genotype, pch = 21, cex = 1.5)
legend(list(x = 0.2, y = 0.3), as.character(levels(prDes$genotype)), pch = 21, pt.bg = c(1, 
    2, 3, 4, 5))
```

Assess how much of the total variance is captured by each principal component:
```{r pca total var}
# Get the subset of PCs that capture the most variance in your predictors
summary(pcs)
```

Select the rotated data from the first `n` components:
```{r pcs2dim}
pcs_2dim = prcomp(sprDat, center = FALSE, scale = FALSE, tol = 0.8)
```

### t-SNE Plots

```{r tsne}
library(Rtsne)
colors = rainbow(length(unique(prDes$grp)))
names(colors) = unique(prDes$grp)
tsne <- Rtsne(unique(t(sprDat)), dims = 2, perplexity = 0.1, verbose = TRUE, max_iter = 100)

plot(tsne$Y, main = "tsne")
text(tsne$Y, labels = prDes$grp, col = colors[prDes$grp])

tsne_p1 <- Rtsne(unique(t(sprDat)), dims = 2, perplexity = 1, verbose = TRUE, max_iter = 100)

plot(tsne_p1$Y, main = "tsne")
text(tsne_p1$Y, labels = prDes$grp, col = colors[prDes$grp])
```

## Deliverables

### Heatmap
> Regenerate the pheatmap clustering plot for the top genes, selected from limma, using clustering distance: correlation, and clustering method: mcquitty.

To construct the heatmap using the top genes, the `topDat` tibble from the `limma` output was used. The parameter `clustering_method` was changed to `mcquitty` and the `clustering_distance_cols` parameter was changed to `correlation`.

```{r heatmap deliverable}
pheatmap(topDat, 
         cluster_rows = FALSE, 
         scale = "none", 
         clustering_method = "mcquitty", 
         clustering_distance_cols = "correlation",
         show_colnames = T, 
         show_rownames = FALSE, 
         main = "Clustering heatmap for GSE70213", 
         annotation = prDes[, c("tissue", "genotype", 
      "grp")])
```


### Dendrogram
> Regenerate the dendrogram on the samples of this heatmap using the `hclust` and `dist` functions.

To construct the dendrogram using `clustering distance: correlation` and `clustering method: mcquitty`, a distance matrix was created using `as.dist()`, with the input as `1-cor(data)`, and then divided by 2 to scale it in order to see the sample labels.

```{r dendrogram deliverable}

# heat map above using clustering distance: correlation; clustering method: mcquitty

pr.dis2 <- as.dist((1-cor(topDat))/2)
pr.hc2 <- hclust(pr.dis2, method = "mcquitty")

plot(pr.hc2, main="mcquitty", xlab="")
```

```{r dendrogram deliverable2, eval = FALSE}

# heat map above using clustering distance: correlation; clustering method: mcquitty

pr.dis <- as.dist(1-cor(topDat))
pr.hc <- hclust(pr.dis, method = "mcquitty")

plot(pr.hc, labels = FALSE, main = "mcquitty", xlab = "")
```

### PCA
> Plot the data for this analyses along PCs 1 and 2 using `ggplot` instead base plotting. Color the points by tissue.

The plot shows that PC2 is tissue, as the points corresponding to the quadriceps cluster towards the top, and the soleus tissue clusters toward the bottom.

```{r pca deliverable}
ggplot(prinComp, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = tissue)) +
  ggtitle("PCA")
```
